---
title: "Death Capstone"
author: "Scott Manusov"
date: "`r Sys.Date()`"
output: html_document
---

library(knitr)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("ggplot2")
install.packages("tidyverse")
install.packages("knitr")
library(dplyr)
library(ggplot2)
library(tidyverse)
library(knitr)

```

## Data Overview (Causes of Death in Worldwide)

This is an academic document reviewing deaths across the world. This will summarize the causes of death from 2017 on with data supplied by the WHO.

```{r data}
print("Hypothesis: Causes of death will shift both overtime within countries, and will have different leading causes between countries")
data <- read.csv('Morticd10_part5.csv')
head(data)
#list of useless columns: Admin1, Subdiv, List, Frmat, IM_Frmat, Deaths 26,
#list of useful columns: Country, Year, Cause, Sex, Deaths1 (all ages), Deaths 2-25, IM_deaths1-4
#Slim down data to data I think I will want to use.  
data <- data %>%
  select(Country, Year, Cause, Sex, Deaths1, Deaths2, Deaths3, Deaths4, Deaths5, Deaths6, Deaths7, Deaths8, Deaths9, Deaths10, Deaths11, Deaths12, Deaths13, Deaths14, Deaths15, Deaths16, Deaths17, Deaths18, Deaths19, Deaths20, Deaths21, Deaths22, Deaths23, Deaths24, Deaths25, Deaths26, IM_Deaths1, IM_Deaths2, IM_Deaths3, IM_Deaths4)
#checking that it worked
head(data)
#convert columns to useful names
data <- data %>%
  rename(
  All_Deaths = Deaths1,
  Age0 = Deaths2,
  Age1 = Deaths3,
  Age2 = Deaths4,
  Age3 = Deaths5,
  Age4 = Deaths6,
  `Age5-9` = Deaths7,
  `Ages10-14` = Deaths8,
  `Ages15-19` = Deaths9,
  `Ages20-24` = Deaths10,
  `Ages25-29` = Deaths11,
  `Ages30-34` = Deaths12,
  `Ages35-39` = Deaths13,
  `Ages40-44` = Deaths14,
  `Ages45-49` = Deaths15,
  `Ages50-54` = Deaths16,
  `Ages55-59` = Deaths17,
  `Ages60-64` = Deaths18,
  `Ages65-69` = Deaths19,
  `Ages70-74` = Deaths20,
  `Ages75-79` = Deaths21,
  `Ages80-84` = Deaths22,
  `Ages85-89` = Deaths23,
  `Ages90-94` = Deaths24,
  `Ages95+` = Deaths25,
  `Age??` = Deaths26,
  `Inf0` = IM_Deaths1,
  `Inf1-6` = IM_Deaths2,
  `Inf7-27` = IM_Deaths3,
  `Inf28-364` = IM_Deaths4
  )
#checking that it worked
head(data)

#Convert to long data format so that I can graph age groups
long_data <- data %>%
  gather(key = "AgeGroup", value = "Deaths", -Cause, -Country, -Sex, -Year)
#Remove the all_Deaths group
ages_data <- long_data %>%
  filter(AgeGroup != "All_Deaths")
#view what table looks like to confirm it worked (*note All_Deaths is a value)
head(ages_data)
  

#First plot what age are people dying by year.  
#Really slow graph (I need to be careful)
#LEARNINGS: Data is bimodal: 
#LEARNINGS: Mode 1 peaks at 85-89 years old.
#LEARNINGS: A very large number of unknown ages (I should filter this out next time)
#LEARNINGS: Infant deaths are much smaller
ggplot(long_data, aes(x= AgeGroup, y = Deaths, fill = as.factor(Year))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Deaths by Age Group and Year",
       x = "Age Group",
       y = "Number of Deaths") +
  scale_fill_discrete(name = "Year") + 
  theme_minimal()

#filter out unknown ages
ages_data <- ages_data %>%
  filter(AgeGroup != "Ages??")
#view what table looks like to confirm it worked (*note All_Deaths is a value)
head(ages_data)

ggplot(long_data, aes(x= AgeGroup, y = Deaths, fill = as.factor(Year))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Deaths by Age Group and Year",
       x = "Age Group",
       y = "Number of Deaths") +
  scale_fill_discrete(name = "Year") + 
  theme_minimal()


#Find just the top 10 causes of deaths worldwide.  
top_10_causes <- ages_data %>%
  group_by(Cause) %>%
  summarize(TotalDeaths = sum(Deaths)) %>%
  top_n(10, TotalDeaths) %>%
  arrange(desc(TotalDeaths))
print(top_10_causes)

#Convert the top 10 types into a more readable function.
#Note reran top_10_causes to confirm it worked
ages_data <- ages_data %>%
  mutate(Cause = case_when(
    Cause == "A97" ~ "Viral Fevers",
    Cause == "B65" ~ "Schistosomiasis",
    Cause == "R94" ~ "Abnormal Symptoms",
    Cause == "B82" ~ "Infectious and Parasitic Dis",
    Cause == "G452" ~ "Meningitis",
    Cause == "A938" ~ "HIV",
    Cause == "A753" ~ "Viral Myocarditis",
    Cause == "A95" ~ "Yellow Fever",
    Cause == "Q73" ~ "Chromosomal Disorder", 
    Cause == "N85" ~ "Genitourinary Dis"
  ))

#Name countries of interest (this will help sort by country as well as help me reduce my data set to make it easier to work with)
ages_data <- ages_data %>%
  mutate(Country = case_when(
    Country == "1010" ~ "Algeria",
    Country == "1100" ~ "Congo",
    Country == "1125" ~ "Egypt",
    Country == "1140" ~ "Ethipia",
    Country == "1180" ~ "Ghana",
    Country == "1190" ~ "Guinea",
    Country == "1220" ~ "Kenya",
    Country == "1240" ~ "Liberia",
    Country == "1310" ~ "Morocco",
    Country == "1370" ~ "Rwanda",
    Country == "1410" ~ "Sierra Leonne",
    Country == "1420" ~ "Somalia",
    Country == "1430" ~ "S. Africa",
    Country == "2025" ~ "Aruba",
    Country == "2030" ~ "Bahamas",
    Country == "2045" ~ "Belize",
    Country == "2050" ~ "Bermuda",
    Country == "2070" ~ "Brazil",
    Country == "2090" ~ "Canada",
    Country == "2190" ~ "El Salvador",
    Country == "2280" ~ "Hondorus",
    Country == "2290" ~ "Jamiaca",
    Country == "2310" ~ "Mexico",
    Country == "2350" ~ "Panama",
    Country == "2370" ~ "Peru",
    Country == "2450" ~ "USA",
    Country == "3025" ~ "Bangladesh",
    Country == "3068" ~ "China",
    Country == "3100" ~ "India",
    Country == "3130" ~ "Iran",
    Country == "3140" ~ "Iraq",
    Country == "3150" ~ "Israel",
    Country == "3160" ~ "Japan",
    Country == "3320" ~ "Qatar",
    Country == "3405" ~ "UAE",
    Country == "4018" ~ "Belarus",
    Country == "4050" ~ "Denmark",
    Country == "4070" ~ "Finland",
    Country == "4080" ~ "France",
    Country == "4084" ~ "Georgia",
    Country == "4085" ~ "Germany",
    Country == "4140" ~ "Greece",
    Country == "4210" ~ "Netherlands",
    Country == "4230" ~ "Poland",
    Country == "4272" ~ "Russia",
    Country == "4290" ~ "Sweden",
    Country == "4300" ~ "Switzerland",
    Country == "4303" ~ "Ukraine",
    Country == "4308" ~ "UK",
    Country == "5070" ~ "Fiji"
      ))
#it would have been smarter to filter first then name them
filtered_countries <- c("Algeria", "Congo", "Egypt", "Ethiopia", "Ghana", "Guinea", "Kenya", "Liberia", "Morocco","Rwanda", "Sierra Leonne", "Somalia", "S. Africa", "Aruba", "Bahamas", "Belize", "Bermuda", "Brazil", "Canada", "El Salvador", "Hondorus", "Jamiaca", "Mexico", "Panama", "Peru", "USA", "Bangladesh", "China", "India", "Iran", "Iraq", "Israel", "Japan", "Qatar", "UAE", "Belarus", "Denmark", "Finland", "France", "Georgia", "Germany", "Greece", "Netherlands", "Poland", "Russia", "Sweden", "Switzerland", "Ukraine", "UK", "Fiji")

country_filtered <- ages_data %>%
  filter(Country %in% filtered_countries)

filtered_causes <- c( "Viral Fevers", "Schistosomiasis", "Abnormal Symptoms", "Infectious and Parasitic Dis", "Meningitis", "HIV", "Viral Myocarditis", "Yellow Fever", "Chromosomal Disorder", "Genitourinary Dis")

cause_filtered <- ages_data %>%
  filter(Cause %in% filtered_causes)

double_filtered <- cause_filtered %>%
  filter(Country %in% filtered_countries)

#graph the the amount of death from the leading causes in my specific countries
ggplot(double_filtered, aes(x= Country, y = Deaths, fill = Cause)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Cause of Death by Country",
       x= "Country",
       y = "Deaths",
       fill = "Cause")


#Trouble shooting 
table_data <- double_filtered %>%
  select(Country, Cause, Deaths) %>%
  group_by(Country, Cause) %>%
  summarize(TotalDeaths = sum(Deaths)) %>%
  ungroup()

print(kable(table_data))
#Graph single country cause of death
Algeria_data <- ages_data %>%
  filter(Country == "Algeria")
leading_causes <- Algeria_data %>%
  summarize(TotalDeaths = sum(Deaths)) %>%
  top_n(5, TotalDeaths) %>%
  arrange(desc(TotalDeaths))
Algeria_filtered <- Algeria_data %>% 
  filter(Cause %in% leading_causes$Cause)
ggplot(Algeria_filtered, aes(x= Cause, y= Deaths, fill = Cause)) +
  geom_bar(stat = "identity") + 
  labs(title = paste("Leading Causes of Death in Algeria"),
       x = "Cause",
       y = "Deaths",
       fille = "Cause") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
#Trouble shooting -found several of the countries in the key are not in the data set 
print(leading_causes)
#update filtered countries:
filtered_countries <- c("Egypt", "S. Africa", "Aruba", "Bermuda", "Brazil", "Canada", "El Salvador", "Mexico", "Panama", "Peru", "USA", "Bangladesh", "Israel", "Japan", "Qatar", "UAE", "Belarus", "Denmark", "Finland", "France", "Georgia", "Germany", "Greece", "Netherlands", "Poland", "Russia", "Sweden", "Switzerland", "Ukraine", "UK")

country_filtered <- ages_data %>%
  filter(Country %in% filtered_countries)

#Retry with country in dataset (note swapped Egypt and Ukraine, since Ukraine is the first country)
Egypt_data <- long_data %>%
  filter(Country == 4303)
#leading_causes <- Egypt_data %>%
 # summarize(TotalDeaths = sum(Deaths)) %>%
  #top_n(5, TotalDeaths) %>%
  #arrange(desc(TotalDeaths))
Egypt_filtered <- Egypt_data %>% 
  #filter(Cause %in% leading_causes$Cause)
ggplot(Egypt_filtered, aes(x= Cause, y= Deaths, fill = Cause)) +
  geom_bar(stat = "identity") + 
  labs(title = paste("Leading Causes of Death in Egypt"),
       x = "Cause",
       y = "Deaths",
       fill = "Cause") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#troubleshooting
print(Egypt_filtered)
unique(long_data$Country)
print(Egypt_data)
head(long_data)

#try again (worked without filtering for top causes):
Egypt_filtered <- Egypt_data

ggplot(Egypt_filtered, aes(x= Cause, y =Deaths, fill = Cause)) + 
  geom_bar(stat = "identity") +
  labs(title = paste("Causes of Death in Egypt"),
       x = "Cause",
       y = "Deaths",
       fill = "Cause") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#adding top causes back in and correcting back to Ukraine:
#just filtering by country
Ukraine_data <- long_data %>%
  filter(Country == 4303)
# finding top causes
top5causes <- Ukraine_data %>%
  group_by(Cause) %>%
  summarize(TotalDeaths = sum(Deaths)) 
print(top5causes) # %>%
  top_n(5, TotalDeaths)
  
str(Ukraine_data)
Ukraine_data <- Ukraine_data[complete.cases(Ukraine_data$Deaths), ]
print(unique(Ukraine_data$Deaths))
#fixing the summarize step:
Ukraine_data <- long_data %>%
  filter(Country == 4303)
# finding top causes
top5causes <- Ukraine_data %>%
  group_by(Cause) %>%
  summarize(TotalDeaths = sum(Deaths)) %>%
  arrange(desc(TotalDeaths)) %>%
  slice_head(n = 5)

print(top5causes) 

#filter for only top causes 
# this isn't needed any more: Ukraine_filtered <- Ukraine_data %>%
  

ggplot(top5causes, aes(x= Cause, y =TotalDeaths, fill = Cause)) + 
  geom_bar(stat = "identity") +
  labs(title = paste("Causes of Death in Ukraine"),
       x = "Cause",
       y = "Deaths",
       fill = "Cause") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

head(Ukraine_data)
head(top5causes)

# finding top causes with years
top5causeyears <- Ukraine_data %>%
  group_by(Cause, Year) %>%
  summarize(TotalDeaths = sum(Deaths)) 

print(top5causeyears) 

top5causeyears <- top5causeyears %>%    
  arrange(desc(TotalDeaths)) %>%
  slice_head(n = 5)

print(top5causeyears)
#new graph years and causes (realized I didn't limit to top 5 causes but still looks good):
ggplot(top5causeyears, aes(x=as.factor(Year), y = TotalDeaths, fill = Cause)) +
  geom_bar(stat = "identity") +
  labs(title = "Ukraine deaths by year and cause",
       x = "Year",
       y = "Deaths",
       fill = "Cause") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Repeat the test, with several other countries:
#Ukraine
Ukraine_data <- long_data %>%
  filter(Country == 4303)

top5causeyears <- Ukraine_data %>%
  filter(AgeGroup == "All_Deaths" ) %>%
  group_by(Cause, Year) %>%
  summarize(TotalDeaths = sum(Deaths)) 

print(top5causeyears) 


ggplot(top5causeyears, aes(x=as.factor(Year), y = TotalDeaths, fill = Cause)) +
  geom_bar(stat = "identity") +
  labs(title = "Ukraine deaths by year and cause",
       x = "Year",
       y = "Deaths",
       fill = "Cause") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Egypt (overwhelmed the key)
Egypt_data <- long_data %>%
  filter(Country == 1125)

top5causeyears <- Egypt_data %>%
  filter(AgeGroup == "All_Deaths" ) %>%
  group_by(Cause, Year) %>%
  summarize(TotalDeaths = sum(Deaths)) 

print(top5causeyears) 


ggplot(top5causeyears, aes(x=as.factor(Year), y = TotalDeaths, fill = Cause)) +
  geom_bar(stat = "identity") +
  labs(title = "Egypt deaths by year and cause",
       x = "Year",
       y = "Deaths",
       fill = "Cause") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#try to create a function that stops the key from getting overwhelmed
country_bar <- function(code, country_name) {
  country_data <- long_data %>%
    filter(Country == code, AgeGroup == "All_Deaths") %>%
    group_by(Cause, Year) %>%
    summarize(TotalDeaths = sum(Deaths)) %>%
    arrange(desc(TotalDeaths)) %>%
    top_n(5)
  
  ggplot(country_data, aes(x = Year, y = TotalDeaths, fill = Cause))+
    geom_bar(stat = "identity") +
    labs(title = paste(country_name, "Deaths by cause and year"),
         x = "Year",
         y = "Deaths",
         fill = " Cause") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

ukraine_plot <- country_bar(4303, "Ukraine")
print(ukraine_plot)
egypt_plot <- country_bar(1125, "Egypt") +
  scale_fill_manual(breaks = country_data$Cause[1:10])
print(egypt_plot)
safrica_plot <- country_bar(1430, "South Africa")
print(safrica_plot)
aruba_plot <- country_bar(4303, "Aruba")
print(aruba_plot)
bermuda_plot <- country_bar(2050, "Bermuda")
print(bermuda_plot)
brazil_plot <- country_bar(2070, "Brazil")
print(brazil_plot)
canada_plot <- country_bar(2090, "Canada")
print(canada_plot)
elsal_plot <- country_bar(2190, "El Salvador")
print(elsal_plot)
mexico_plot <- country_bar(2310, "Mexico")
print(mexico_plot)
panama_plot <- country_bar(2350, "Panama")
print(panama_plot)
peru_plot <- country_bar(2370, "Peru")
print(peru_plot)
usa_plot <- country_bar(2450, "USA")
print(usa_plot)
iraq_plot <- country_bar(31400, "Iraq")
print(iraq_plot)
israel_plot <- country_bar(3150, "Israel")
print(israel_plot)
japan_plot <- country_bar(3160, "Japan")
print(peru_plot)
qatar_plot <- country_bar(3320, "Qatar")
print(qatar_plot)
uae_plot <- country_bar(3405, "UAE")
print(uae_plot)
belarus_plot <- country_bar(4018, "Belarus")
print(belarus_plot)
denmark_plot <- country_bar(4050, "Denmark")
print(denmark_plot)
finland_plot <- country_bar(4070, "Finland")
print(finland_plot)
france_plot <- country_bar(4080, "France")
print(france_plot)
georgia_plot <- country_bar(4084, "Georgia")
print(georgia_plot)
germany_plot <- country_bar(4085, "Germany")
print(germany_plot)
germany_plot <- country_bar(4140, "Greece")
print(germany_plot)
germany_plot <- country_bar(4210, "Netherlands")
print(germany_plot)
germany_plot <- country_bar(4230, "Poland")
print(germany_plot)
germany_plot <- country_bar(4272, "Russia")
print(germany_plot)
germany_plot <- country_bar(4290, "Sweden")
print(germany_plot)
germany_plot <- country_bar(4300, "Switzerland")
print(germany_plot)
germany_plot <- country_bar(4303, "Ukraine")
print(germany_plot)
germany_plot <- country_bar(4308, "UK")
print(germany_plot)

print("Hypothesis: Causes of death will shift both overtime within countries, and will have different leading causes between countries")
print("Conclusion 1: Causes of death from 2017 - 2019 did not shift within most if any country\n Conclusion 2: Countries not only have different causes, but use different naming and sorting systems, making it difficult and time taking to determine the full extent of the differences")

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
